---
title: "Gov 52 Final Replication Report"
subtitle: "The Personality of the Politically Ambitious (Dynes, Hassell, and Miles)"
output: pdf_document
name: Ella Michaels
---
_Code for this project can be found at [CODE CODE CODE]_

```{r setup, messages = FALSE, warnings = FALSE, include=FALSE, header = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library("effects")
library("foreign")
library("gmodels")
library("gplots")
library("nlme")
library("nnet")
library("stargazer")
library(MASS)
library(rockchalk)
library(pscl)
library(car)
library(DescTools)
library(psych)
library(MNP)
library(knitr)

# Loading/cleaning genpop data
d1 <- read.dta("data/genpop replication.dta") %>%
  mutate(race = relevel(race, ref = "White"),
         dideo = relevel(dideo, ref = "Moderate"),
         PartyID = relevel(PartyID, ref = "Independent"),
         educ = relevel(educ, ref = "Some College"),
         Age = Recode(Age, "-936 = '18"),
         Age = as.numeric(Age) + 17) # ages had 17 subtracted for some reason 

##Logit Models
mod.40 = multinom(Q23_1~Extra+Open+consc+Agree+emotstab+educ+income+race+dideo+PartyID+gender+Age, data=d1) # not used
mod.50 = polr(Q23_1~Extra+Open+consc+Agree+emotstab+educ+income+race+dideo+PartyID+gender+Age, data=d1)

# No controls
mod.140 = multinom(Q23_1~Extra+Open+consc+Agree+emotstab, data=d1) # not used
mod.150 = polr(Q23_1~Extra+Open+consc+Agree+emotstab, data=d1)

```

# Overview

Historically, political ambition has been considered a product of external environments. However, as part of a broader movement emphasizing the role of individual personality in political outcomes and trajectories, Dynes et al. offer insight on how personality shapes political ambition. Their research helps demonstrate how personality affects both "nascent" and "progressive" political ambition. That is, a someone's interest in an initial run for public office, and a current politician's interest in running for higher office. 

They conduct two separate studies to test the effect of personality on nascent and progressive ambition, respectively. The first, on nascent political ambition, used a nationally representative survey of US adults. Respondents were asked to fill out a personality scale, used to measure where they fell on each facet of the five-factor model of personality. They also indicated their "attitude toward running for office in the future." Results suggested that those with higher levels of extroversion and openness were more likely to consider running for office, while those who were more agreeable and conscientious were less likely.

The second study, this time on progressive ambition, used a nationally representative survey of local US politicians. Respondents were again asked to fill out a personality scale and answer a number of other questions about their future political plans. Those with higher levels of extroversion, openness, _and_ emotional stability were more likely to be progressively ambitious. Higher levels of agreeableness were negatively associated with progressive ambition, but this time around conscientiousness was less of a factor. Notably, personality was generally a weaker predictor of progressive ambition than nascent ambition.

# Replication Process

I replicated the primary analyses and figures for both parts of the study (on nascent and progressive ambition, respectively). All models included use ordered logistic regression, since all of the dependent variables on ambition are ordered. Model versions using multinomial logistic regression were included in the authors' appendix, but were not summarized in the primary paper and yield consistent results. In each table, the output for the model without controls (i.e. only five factor personality traits are tested) is represented in the first column. The output for the model in the second column includes controls.

The authors' provided code included a variety of model formulations, many of which were not included in the main study. To create the tables, I identified those used in the final paper, and curated and formatted the appropriate stargazer outputs. For the figures, I cleaned and compiled the relevant `effects` outputs and used ggplot to create the graphics, which were not included in the authors' provided code.

### Study 1: Nascent Political Ambition

The response variable in study one was attitude towards running for political office among people who had never been politicians. Survey participants ranked their attitude from "absolutely no interest," to "open to the possibility," to "actively considering." 

Figure 1 illustrates the predicted probabilities of being "open to the possibility" of running for political office at various levels of different personality traits, based on the controlled version of the ordered logit model in Table 1. These probabilities were generated using the effects package.

```{r Table 1, results='asis', warning=FALSE, message=FALSE, header = FALSE}


## TABLE 1 OUTPUT - ordered logit used
stargazer(mod.150,mod.50, type='latex', style='default',
          column.labels = c("No controls", "Demographic controls"),
          title = "Nascent Political Ambition",
          dep.var.caption = "Attractiveness of Elected Office",
          dep.var.labels = "",
          header = FALSE,
          font.size = "scriptsize",
          covariate.labels = c("Extroversion", "Openness to experience", "Agreeableness",
                               "Conscientiousness", "Emotional stability", 
                               "Education: Less than high school", "Education: High school",
                               "Education: Bachelor's degree", "Education: Graduate degree",
                               "Income", "Race: Afican American", "Race: Asian", 
                               "Race: Native American", "Race: Hispanic", 
                               "Race: Multiracial", "Ideology: Liberal", 
                               "Ideology: Conservative", "Party ID: Democrat", 
                               "PartyID: Republican", "Gender: Female", "Age (in years)",
                               "Observations"))

```

```{r Figure 1 cleaning, warning=FALSE, message=FALSE}

# Original replication code used model 51, but the models in tables are 50 and 150, AND the figure caption says it used ordered logit in the subtitle (51 is multinomial) so model 50 (with controls) is used here. Neither 50 nor 51 yield the exact results in the paper - not sure why. 
# Added correct leveling, selected relevant columns

effagree <- as.data.frame(effect(c('Agree'), mod=mod.50, xlevels=list(Agree=c(0,0.5, 1, 1.5, 2,2.5,3)))) %>%
  dplyr::select("Agree", "prob.I.am.open.to.the.possibility.of.holding.elective.office.in.the.future.") %>%
  rename("possible" = "prob.I.am.open.to.the.possibility.of.holding.elective.office.in.the.future.",
         "score" = "Agree") %>%
  mutate(type = "Agreeableness")

effcons = as.data.frame(effect(c('consc'), mod=mod.50, xlevels=list(consc=c(0,0.5, 1, 1.5, 2,2.5,3)))) %>%
  dplyr::select("consc", "prob.I.am.open.to.the.possibility.of.holding.elective.office.in.the.future.") %>%
  rename("possible" = "prob.I.am.open.to.the.possibility.of.holding.elective.office.in.the.future.",
         "score" = "consc") %>%
  mutate(type = "Conscientiousness")

effextra = as.data.frame(effect(c('Extra'), mod=mod.50, xlevels=list(Extra=c(0,0.5, 1, 1.5, 2,2.5,3)))) %>%
  dplyr::select("Extra", "prob.I.am.open.to.the.possibility.of.holding.elective.office.in.the.future.") %>%
  rename("possible" = "prob.I.am.open.to.the.possibility.of.holding.elective.office.in.the.future.",
         "score" = "Extra") %>%
  mutate(type = "Extroversion")

effopen = as.data.frame(effect(c('Open'), mod=mod.50, xlevels=list(Open=c(0,0.5, 1, 1.5, 2,2.5,3)))) %>%
  dplyr::select("Open", "prob.I.am.open.to.the.possibility.of.holding.elective.office.in.the.future.") %>%
  rename("possible" = "prob.I.am.open.to.the.possibility.of.holding.elective.office.in.the.future.",
         "score" = "Open") %>%
  mutate(type = "Openness")

effneuro = as.data.frame(effect(c('emotstab'), mod=mod.50, xlevels=list(emotstab=c(0,0.5, 1, 1.5, 2,2.5,3)))) %>%
  dplyr::select("emotstab", "prob.I.am.open.to.the.possibility.of.holding.elective.office.in.the.future.") %>%
  rename("possible" = "prob.I.am.open.to.the.possibility.of.holding.elective.office.in.the.future.",
         "score" = "emotstab") %>%
  mutate(type = "Emotional Stability")

# combining
fig1_data <- rbind(effagree, effcons, effextra, effopen, effneuro)

```



```{r Figure 1, warning=FALSE, message=FALSE}

fig1_data %>%
  ggplot(aes(x = score, y = possible, shape = type)) +
  geom_line() +
  geom_point() +
  xlim(0,3) +
  ylim(0,.4) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line( size=.2, color="grey" ),
        panel.grid.minor.y = element_line( size=.2, color="grey" ),
        panel.grid.minor.x = element_blank(),
        axis.line.y.left = element_line(size=.2, color="grey")) +
  labs(title = "Figure 1: Open to the Possibility of Seeking Elective Office",
       x = "Big Five Personality Score",
       y = "Possibility (Open to Possibility)",
       linetype = "",
       shape = "") +
  scale_x_continuous(breaks = c(0, 0.5, 1, 1.5, 2, 2.5,3)) 

ggsave("figure1.jpg", width = 13)

```

## Study 2: Progressive Political Ambition

Study 2 relied on survey data from a nationally representative sample of local elected officials. There were two response variables under review. The first, under observation in Table 2 and Figure 2, was a four-level survey result gauging interest in running for elected office again from 1) "something I would absolutely never do" to 4) "I definitely would like to".

Figure 2 illustrates the predicted probabilities of response level 3) "It is something I might undertake if the opportunity presented
itself", based on the controlled version of the ordered logit model in Table 2. 

Table three examines the third overall response variable, namely the likelihood of running for higher political offices (i.e. from local to state to national). Recall that the politicians surveyed were all local officials. Agreeableness was negatively associated, and openness to experience was positively associated, with desires to seek higher office.


```{r AMOS cleaning, warning=FALSE, message=FALSE}

d2 <- read.dta("data/AMOS replication.dta") %>%
  mutate(runhigher = factor(progamb_runhigher),
         progamb = as.numeric(runhigher))
d2$progamb_winlegis_1[d2$progamb_winlegis_1==-99] <- NA


```


 

```{r Table 2, results='asis', warning=FALSE, message=FALSE}

# ordered logit, no controls (col 1, table 2)
mod.71=polr(runhigher~extra4pt+ open4pt+agree4pt+ consc4pt+ stable4pt, data=d2)

# ordered logit, controls (col 2, table 2)
mod.71c <- polr(runhigher~extra4pt+ open4pt+agree4pt+ consc4pt+ stable4pt+ closevote+ tenure+ progamb_current+termlimits+ partisanelect+ progamb_similar_1+ progamb_winlegis_1 + gender, data=d2)
# coefficients slightly different - appears something is different about the data because the observation numbers are slightly off. Tried it with their exact code so not sure what's going on here.

mod.72 = polr(runhigher~extra4pt+ open4pt+agree4pt+ consc4pt+ stable4pt+ closevote+ tenure+ progamb_current+termlimits+ partisanelect+ tenure+ progamb_similar_1+ progamb_winlegis_1, data=d2)

# Table 2
stargazer(mod.71, mod.71c, type = "latex",
          header = FALSE,
          column.labels = c("No controls", "Demographic controls"),
          dep.var.caption = "Progressive Political Ambition",
          dep.var.labels = "",
          title = "Progressive Political Ambition",
          covariate.labels = c("Extroversion", "Openness to experience", "Agreeableness",
                               "Conscientiousness", "Emotional stability", 
                               "Won previous election by 5% pts. or less", "Years in office",
                               "Anticipated length in current office", 
                               "Term limits exist for current office", "Partisan elections",
                               "Probability current seat filled by similar candidate",
                               "Probability similar candidate could win legislative seat",
                               "Gender: Female", "Observations"))


```


```{r Fig 2 data cleaning, warning=FALSE, message=FALSE, include = FALSE}

# they use mod 72, but that's not what's displayed (same as fig 1 in this regard)

mod.72.o.ef = as.data.frame(effect(c('open4pt'),mod=mod.71c, xlevels=list(open4pt=c(0,0.5, 1, 1.5, 2,2.5,3)))) %>%
  dplyr::select("open4pt", "prob.X2") %>%
  rename("maybe" = "prob.X2",
         "score" = "open4pt") %>%
  mutate(type = "Openness")

mod.72.c.ef = as.data.frame(effect(c('consc4pt'),mod=mod.71c, xlevels=list(open4pt=c(0,0.5, 1, 1.5, 2,2.5,3)))) %>%
  dplyr::select("consc4pt", "prob.X2") %>%
  rename("maybe" = "prob.X2",
         "score" = "consc4pt") %>%
  mutate(type = "Conscientiousness")
                                   
mod.72.e.ef = as.data.frame(effect(c('extra4pt'),mod=mod.71c, xlevels=list(open4pt=c(0,0.5, 1, 1.5, 2,2.5,3)))) %>%
  dplyr::select("extra4pt", "prob.X2") %>%
  rename("maybe" = "prob.X2",
         "score" = "extra4pt") %>%
  mutate(type = "Extroversion")

mod.72.a.ef = as.data.frame(effect(c('agree4pt'),mod=mod.71c, xlevels=list(open4pt=c(0,0.5, 1, 1.5, 2,2.5,3)))) %>%
  dplyr::select("agree4pt", "prob.X2") %>%
  rename("maybe" = "prob.X2",
         "score" = "agree4pt") %>%
  mutate(type = "Agreeableness")

mod.72.n.ef = as.data.frame(effect(c('stable4pt'),mod=mod.71c, xlevels=list(open4pt=c(0,0.5, 1, 1.5, 2,2.5,3)))) %>%
  dplyr::select("stable4pt", "prob.X2") %>%
  rename("maybe" = "prob.X2",
         "score" = "stable4pt") %>%
  mutate(type = "Emotional Stability")

fig2_data <- rbind(mod.72.o.ef, mod.72.c.ef, mod.72.e.ef, mod.72.a.ef, mod.72.n.ef)

```


```{r Figure 2, warning=FALSE, message=FALSE}

fig2_data %>%
  ggplot(aes(x = score, y = maybe, shape = type)) +
  geom_line() +
  geom_point() +
  xlim(0,3) +
  ylim(0,.4) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line( size=.2, color="grey" ),
        panel.grid.minor.y = element_line( size=.2, color="grey" ),
        panel.grid.minor.x = element_blank(),
        axis.line.y.left = element_line(size=.2, color="grey")) +
  labs(title = "Figure 2: Open to the Possibility of Seeking Higher Office",
       x = "Big Five Personality Score",
       y = "Probability (If the Opportunity Presented Itself)",
       linetype = "",
       shape = "") +
  scale_x_continuous(breaks = c(0, 0.5, 1, 1.5, 2, 2.5,3)) 

ggsave("figure2.jpg", width = 13)

```

```{r Table 3 prep, results='asis', warning=FALSE, message=FALSE, include = FALSE}

# Ordered logit
mod.100 = polr(office~extra4pt+ open4pt+agree4pt+ consc4pt+ stable4pt, data=d2)
mod.101 = polr(office~extra4pt+ open4pt+agree4pt+ consc4pt+ stable4pt+ closevote+ tenure+ progamb_current+termlimits+ partisanelect+ tenure+ progamb_similar_1+ progamb_winlegis_1+gender, data=d2)

# Multinomial (not used)
mod.111 = multinom(office~extra4pt+ open4pt+agree4pt+ consc4pt+ stable4pt+ closevote+ tenure+ progamb_current+termlimits+ partisanelect+ tenure+ progamb_similar_1+ progamb_winlegis_1+gender, data=d2)

```


```{r Table 3, results='asis', warning=FALSE, message=FALSE}
# Table 3
stargazer(mod.100,mod.101, type='latex', style='default',
          header = FALSE,
          column.labels = c("No controls", "Demographic controls"),
          dep.var.caption = "Appeal of Higher Office",
          dep.var.labels = "",
          title = "Progressive Political Ambition",
          covariate.labels = c("Extroversion", "Openness to experience", "Agreeableness",
                               "Conscientiousness", "Emotional stability", 
                               "Won previous election by 5% pts. or less", "Years in office",
                               "Anticipated length in current office", 
                               "Term limits exist for current office", "Partisan elections",
                               "Probability current seat filled by similar candidate",
                               "Probability similar candidate could win legislative seat",
                               "Gender: Female", "Observations"))

```


_Inconsistency Note:_ While I was able to exactly recreate the results of most of the models under review, my outputs for the progressive political ambition models that included controls did not exactly match those in the original paper. The recorded sample sizes in my model outputs were also smaller. I suspect that the data available for replication may have included more NAs in some of the control columns that the actual set used for the study. While there are slight differences, the results are still remarkably similar. The figures, which used controlled results, were also slightly affected, but again, the broad takeaways still held.
